sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-09 doc/paper.Rmd
Edits

<!-- or areal/grid data with centroids instead of CAR/SAR models -->
<!-- add CAR/SAR to table -->
sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
Geostatistical data simply refers to data observed at specific spatial coordinates. 
The data can be collected across discrete points in time.
Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR).
sdmTMB can also fit models with areal data if each polygon has an associated centroid.
<!-- (observations are treated as replicates). -->
A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., range).

@@

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].

@@

# An introductory example

@@

# Spatially varying coefficients from citizen-science count data

@@

# Additional functionality

## Model validation and selection

@@

## Parameter priors/penalties

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations [@simpson2017; @breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.

## Derived quantities of interest

The results from an sdmTMB model can be used to generate various quantities of interest.
For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`; Fig. \@ref(fig:pcod-fig)E) given a user supplied grid.
Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.

@@

## Speed

@@

## Functionality

>>>>

<!-- or areal/grid data with centroids instead of CAR/SAR models -->
<!-- add CAR/SAR to table -->
sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
Geostatistical data simply refers to data observed at specific spatial coordinates. 
The data can be collected across discrete points in time.
Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR).
sdmTMB can also fit models with areal data if each polygon has an associated centroid.
<!-- (observations are treated as replicates). -->
A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., range---the distance at which points are effectively independent).

@@

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template calculates the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors [@simpson2017] (`?pc_matern`) can constrain the spatial range and variance parameters.

The results from an sdmTMB model can be used to generate various quantities of interest.
For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`) given a user-supplied grid.
Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.

@@

# Introductory example with fish survey biomass

@@

# Spatially varying coefficient example with citizen science data

@@

# Model validation and selection

@@

@@

@@
